---
title: "FULL UGLY MARKDOWN"
author: "Adrian Janucik"
date: "27/08/2021"
output: html_document

---

```{r Setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
 # Enable code exectuion log globally
knitr::opts_chunk$set(message = FALSE) # Disable log messages
knitr::opts_chunk$set(warning = FALSE) # Disable warning logs
```
## Quality Control And Differentiall Gene Expression Results

#### This Is A Report Of Quality Control Analyses Ran On The Normalized Raw Counts Using DESeq2.

## PCA Plots

```{r startup_and_2D_PCA_initialization, echo=FALSE, fig.height=6, fig.width=9.5}
suppressPackageStartupMessages(library(BiocManager)) # Before deploying remember to add BiocManager to library sources
options(repos = BiocManager::repositories())

options(ggrepel.max.overlaps = Inf) # Allow for overlaps on plots

# Dependencies ------------------------------------------------------------
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(InteractiveComplexHeatmap))
suppressPackageStartupMessages(library(dendextend))
suppressPackageStartupMessages(library(heatmaply))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(hrbrthemes))
suppressPackageStartupMessages(library(ggthemes))

# Setup -------------------------------------------------------------------
meta <- read_csv('sampleTable2.csv') # Read table with meta contents

# Read in DESeq2 normalized count matrix (median of ratios)
deseq_normalized_counts <-
  read_csv('GTF_FIXED_deseq_featureCounts_normalised_counts.csv') %>%
  column_to_rownames(var = 'id') %>%
  select(-c(1))

# Temporary solution | Double the matrix
deseq_normalized_counts2 <- deseq_normalized_counts

# Read in DE results
dge <- read_csv('GTF_FIXED_deseq_featureCounts_dge.csv') %>%
  select(-1)

# PCA on DESeq2 normalized counts -----------------------------------------

# Perform PCA focusing only on first three dimensions for 3D plot
deseq_normalized_counts_pca <- deseq_normalized_counts %>%
  t() %>%
  prcomp(rank. = 3)

# Perform PCA focusing only on first two dimensions for 2D plot
deseq_normalized_counts_pca_2d <- deseq_normalized_counts %>%
  t() %>%
  prcomp(rank. = 2)

# Temporary solution | Assign meta information to variables
condition = meta$condition
sample = meta$samples

# Call summary on the PCA to acquire explained variance
deseq_normalized_counts_pca_2d_summary <-
  summary(deseq_normalized_counts_pca_2d)

# Create label for PCA showing % of explained variance
tot_explained_variance_ratio_2d <-
  summary(deseq_normalized_counts_pca_2d)$importance[2, 1:2]
tot_explained_variance_ratio_2d <-
  100 * sum(tot_explained_variance_ratio_2d)

```

##### 2D PCA on DESeq2's Normalized Counts

##### Total Explained Variance = `r tot_explained_variance_ratio_2d`%

```{r plot_2D_PCA_&_create_3D_label, echo=F, fig.height=7, fig.width=9}
library(ggplot2)
pca2d_1 <- ggplot(data.frame(deseq_normalized_counts_pca_2d$x), aes(PC1, PC2, color = condition, label = sample)) +
  geom_jitter(size = 3) +
  theme_hc()

pca2d_1 <- ggplotly(pca2d_1)
pca2d_1 <- layout(pca2d_1, margin=list(t = 75))
pca2d_1
# Call summary on the PCA to acquire explained variance
deseq_normalized_counts_pca_summary <- summary(deseq_normalized_counts_pca)

# Create label for PCA showing % of explained variance
tot_explained_variance_ratio <- summary(deseq_normalized_counts_pca)$importance[2,1:3]
tot_explained_variance_ratio <- 100 * sum(tot_explained_variance_ratio)

# Function calling PCA | for specifics check the link in the section above
```
##### 3D PCA on DESeq2's Normalized Counts

##### Total Explained Variance = `r tot_explained_variance_ratio`%
```{r fig1, echo = F, fig.height = 7, fig.width=9}
  (fig1 <-
    plot_ly(
      data.frame(deseq_normalized_counts_pca$x),
      text = ~rownames(deseq_normalized_counts_pca$x),
      x = ~ PC1,
      y = ~ PC2,
      z = ~ PC3,
      color = ~ meta$condition,
      colors = c('#BF382A', '#0C4B8E')
    ) %>%
    add_markers() %>%
    layout(scene = list(
      xaxis = list(title = paste("PC1 -", round(deseq_normalized_counts_pca_summary$importance[2, 1], 3))),
      yaxis = list(title = paste("PC2 -", round(deseq_normalized_counts_pca_summary$importance[2, 2], 3))),
      zaxis = list(title = paste("PC3 -", round(deseq_normalized_counts_pca_summary$importance[2, 3], 3)))
    )))

# PCA on DESeq2 normalized counts - ONLY DEGs -----------------------------
# https://plotly.com/r/3d-scatter-plots/

# Filter DE results for DEGs
degs <- dge %>% 
  filter(padj < 0.1) %>% 
  filter(log2FoldChange > 0.5 | log2FoldChange < -0.5)
####
deseq_normalized_counts <- deseq_normalized_counts[rownames(deseq_normalized_counts) %in% degs$id,]
deseq_normalized_counts_pca_2d <- deseq_normalized_counts %>%
  t() %>%
  prcomp(rank. = 2)

condition = meta$condition
sample = meta$samples

# Call summary on the PCA to acquire explained variance
deseq_normalized_counts_pca_2d_summary <- summary(deseq_normalized_counts_pca_2d)

# Create label for PCA showing % of explained variance
tot_explained_variance_ratio_2d <- summary(deseq_normalized_counts_pca_2d)$importance[2,1:2]
tot_explained_variance_ratio_2d <- 100 * sum(tot_explained_variance_ratio_2d)


library(ggplot2)
pca2d_2 <- ggplot(data.frame(deseq_normalized_counts_pca_2d$x), aes(PC1, PC2, color = condition, label = sample)) +
  geom_jitter(size = 3) +
  theme_hc()
pca2d_2 <- ggplotly(pca2d_2)
pca2d_2 <- layout(pca2d_2, margin=list(t = 75))
```
##### 2D PCA on DESeq2's Normalized Counts

##### Total Explained Variance = `r tot_explained_variance_ratio_2d`% 
```{r pca2d_2, echo = F, fig.height = 7, fig.width=9}
pca2d_2

####
# Perform PCA focusing only on first three dimensions
deseq_normalized_counts_pca_degs <- deseq_normalized_counts[rownames(deseq_normalized_counts) %in% degs$id,] %>%
  t() %>%
  prcomp(rank. = 3)

# Call summary on the PCA to acquire explained variance
deseq_normalized_counts_pca_degs_summary <- summary(deseq_normalized_counts_pca_degs)

# Create label for PCA showing % of explained variance
tot_explained_variance_ratio <- summary(deseq_normalized_counts_pca_degs)$importance[2,1:3]
tot_explained_variance_ratio <- 100 * sum(tot_explained_variance_ratio)

# Function calling PCA | for specifics check the link in the section above
```
##### 3D PCA on DESeq2's Normalized Counts

##### Total Explained Variance = `r tot_explained_variance_ratio`%
```{r fig2, echo = F, fig.height = 7, fig.width=9}
  (fig2 <-
    plot_ly(
      data.frame(deseq_normalized_counts_pca_degs$x),
      text = ~rownames(deseq_normalized_counts_pca_degs$x),
      x = ~ PC1,
      y = ~ PC2,
      z = ~ PC3,
      color = ~ meta$condition,
      colors = c('#BF382A', '#0C4B8E')
    ) %>%
    add_markers() %>%
    layout(scene = list(
      xaxis = list(title = paste("PC1 -", round(deseq_normalized_counts_pca_degs_summary$importance[2, 1], 3))),
      yaxis = list(title = paste("PC2 -", round(deseq_normalized_counts_pca_degs_summary$importance[2, 2], 3))),
      zaxis = list(title = paste("PC3 -", round(deseq_normalized_counts_pca_degs_summary$importance[2, 3], 3)))
    )))

```

## Dendrograms representing sample clustering

##### Dendrogram on DESeq2's Normalized Counts

```{r pressure, echo=FALSE, fig.height = 7, fig.width=9}
# Dendrogram on DESeq2 normalized counts  ---------------------------------
deseq_normalized_counts <- deseq_normalized_counts2
# Specify factors and their levels
meta$condition <-
  factor(meta$condition, levels = c('NonAsthma', 'Asthma'))

# Create distance matrix
d_deseq_normalized_counts <- dist(t(deseq_normalized_counts))
hc_deseq_normalized_counts <- hclust(d_deseq_normalized_counts, method = "complete")
conditions <- rev(levels(meta$condition))

dend <- as.dendrogram(hc_deseq_normalized_counts)

# Color the branches based on the clusters:
dend <- color_branches(dend, k = 2, col = c('#BF382A', '#0C4B8E'))#, groupLabels=conditions)

# Manually match the labels, as much as possible, to the real classification of the flowers:
labels_colors(dend) <-
  c('#BF382A', '#0C4B8E')[sort_levels_values(as.numeric(meta$condition)[order.dendrogram(dend)])]

# We shall add the flower type to the labels:
labels(dend) <-
  paste(as.character(meta$condition)[order.dendrogram(dend)],
        "(", labels(dend), ")",
        sep = "")

# We hang the dendrogram a bit:
dend <- hang.dendrogram(dend, hang_height = 0.1)

# reduce the size of the labels:
dend <- set(dend, "labels_cex", 0.6)

# And plot:
par(mar = c(2, 2, 2, 14) + 0.1, xpd = NA)
plot(dend,
     horiz =  TRUE,
     nodePar = list(cex = .007))
legend("topleft",
       legend = conditions,
       fill = c('#BF382A', '#0C4B8E'))

meta$colorCondition <-
  ifelse(as.character(meta$condition) == 'Asthma', '#BF382A', '#0C4B8E')
colored_bars(
  colors = meta$colorCondition,
  dend = dend,
  rowLabels = "",
  add = TRUE,
  horiz = TRUE
)
```

#### Dendrogram on DESeq2 normalized counts with only DEGs

```{r dendrogram, echo=FALSE, fig.height = 7, fig.width=9}
# Dendrogram on DESeq2 normalized counts - ONLY DEGs ----------------------
meta$condition <-
  factor(meta$condition, levels = c('NonAsthma', 'Asthma'))

deseq_normalized_counts_degs <- deseq_normalized_counts[rownames(deseq_normalized_counts) %in% degs$id,]

d_deseq_normalized_counts_degs <- dist(t(deseq_normalized_counts_degs))
hc_deseq_normalized_counts_degs <- hclust(d_deseq_normalized_counts_degs, method = "complete")
conditions <- rev(levels(meta$condition))

# Create dendrogram object
dend_degs <- as.dendrogram(hc_deseq_normalized_counts_degs)

# Color the branches based on the clusters:
dend_degs <- color_branches(dend_degs, k = 2, col = c('#BF382A', '#0C4B8E'))#, groupLabels=conditions)

# Manually match the labels, as much as possible, to the real classification of the flowers:
labels_colors(dend_degs) <-
  c('#BF382A', '#0C4B8E')[sort_levels_values(as.numeric(meta$condition)[order.dendrogram(dend_degs)])]

# We shall add the flower type to the labels:
labels(dend_degs) <-
  paste(as.character(meta$condition)[order.dendrogram(dend_degs)],
        "(", labels(dend_degs), ")",
        sep = "")

# We hang the dendrogram a bit:
dend_degs <- hang.dendrogram(dend_degs, hang_height = 0.1)

# reduce the size of the labels:
dend_degs <- set(dend_degs, "labels_cex", 0.6)

# And plot:
par(mar = c(2, 2, 2, 14) + 0.1, xpd = NA)
plot(dend_degs,
     horiz =  TRUE,
     nodePar = list(cex = .007))
legend("topleft",
       legend = conditions,
       fill = c('#BF382A', '#0C4B8E'))
meta$colorCondition <-
  ifelse(as.character(meta$condition) == 'Asthma', '#BF382A', '#0C4B8E')
colored_bars(
  colors = meta$colorCondition,
  dend = dend_degs,
  rowLabels = "",
  add = TRUE,
  horiz = TRUE
)

```

## Correlation between samples performed on Euclidean distance with the use of DE genes

```{r heatmap, echo = FALSE, fig.height = 6, fig.width = 10}
# Heatmap -----------------------------------------------------------------


gradient_col <- ggplot2::scale_fill_gradient2(
  low = "blue", high = "red", 
  midpoint = 0, limits = c(-1, 1)
)

side_annot <- meta$condition %>% 
  sort(decreasing = TRUE)

side_annot_top <- meta$condition %>% 
  sort(decreasing = FALSE)

deseq_normalized_counts_distance <- deseq_normalized_counts_degs

deseq_normalized_counts_dist_mat_cor <- dist(t(deseq_normalized_counts_distance),
                                             diag = TRUE,
                                             upper = TRUE) %>% 
  as.matrix() %>% 
  cor()

colnames(deseq_normalized_counts_dist_mat_cor) <- paste(meta$condition, meta$samples, sep = "-")
rownames(deseq_normalized_counts_dist_mat_cor) <- colnames(deseq_normalized_counts_dist_mat_cor)
sorted_deseq_normalized_counts_dist_mat <- deseq_normalized_counts_dist_mat_cor[order(rownames(deseq_normalized_counts_dist_mat_cor)), order(colnames(deseq_normalized_counts_dist_mat_cor), decreasing = TRUE)]

dit_cor_height = nrow(sorted_deseq_normalized_counts_dist_mat)*0.25
```

```{r heatmap_dit_cor_1, echo = F, fig.height=dit_cor_height, fig.width=10}
heatmap_dit_cor <- heatmaply(
  sorted_deseq_normalized_counts_dist_mat,
  scale_fill_gradient_fun = gradient_col,
  col_side_colors = data.frame("Condition" = side_annot_top),
  row_side_colors = data.frame("Condition" = side_annot),
  limits = c(-1,1),
  Rowv = FALSE,
  Colv = FALSE,
  showticklabels = TRUE,
  row_side_palette = colorRampPalette(c('#BF382A', '#0C4B8E'), bias=1),
  col_side_palette = colorRampPalette(c('#BF382A', '#0C4B8E'), bias=1),
  key.title = 'corr',
  height = nrow(sorted_deseq_normalized_counts_dist_mat)*0.7)

s <- subplot(heatmap_dit_cor, margin = .01, titleY = TRUE) %>% 
  config(displayModeBar = F)
s$x$data[[1]]$showlegend <- FALSE
s$x$data[[11]]$showlegend <- FALSE
s$x$layout$annotations[[2]]$text <- 'Condition'
s$x$layout$annotations[[2]]$y <- 0.8
s$x$layout$annotations[[2]]$x <- 1.026

s


```


## Volcano Plots


```{r vlocano, echo = FALSE, fig.height = 13, fig.width=18}
# Volcano -----------------------------------------------------------------
keyvals <-
  ifelse(dge$id %in% c(""),
         'red',
         ifelse((
           dge$log2FoldChange > 0.5 |
             dge$log2FoldChange < -0.5
         ) & dge$padj < 0.1,
         'orange',
         ifelse(dge$padj > 0.1, '#143D59', '#143D59')
         ))

keyvals[is.na(keyvals)] <- '#143D59'
#names(keyvals)[keyvals == 'red'] <- 'Commonly Regulated'
names(keyvals)[keyvals == '#143D59'] <-
  'Not Statistically Significant'
names(keyvals)[keyvals == 'orange'] <- 'Log2FC & padj'

v1 <- EnhancedVolcano::EnhancedVolcano(
  dge,
  lab = dge$id,
  x = 'log2FoldChange',
  y = 'padj',
  title = 'Asthma vs Non-Asthma',
  pCutoff = 0.1,
  FCcutoff = 0.5,
  pointSize = 3.5,
  labSize = 4.5,
  colCustom = keyvals,
  #col=c('#143D59', '#143D59', '#143D59', 'orange'),
  labCol = 'black',
  labFace = 'bold',
  boxedLabels = T,
  colAlpha = 3.5 / 5,
  legendPosition = 'top',
  legendLabSize = 18,
  legendIconSize = 5.0,
  drawConnectors = T,
  widthConnectors = 1.5,
  colConnectors = 'black',
  gridlines.major = T,
  gridlines.minor = FALSE,
  border = 'full',
  borderWidth = 1,
  borderColour = 'black',
  #xlim = c(-4,4),
  #ylim = c(0,11)
)

v2 <- EnhancedVolcano::EnhancedVolcano(
  dge,
  lab = dge$id,
  x = 'log2FoldChange',
  y = 'padj',
  title = 'Asthma vs Non-Asthma',
  pCutoff = 0.1,
  FCcutoff = 0.5,
  pointSize = 3.5,
  labSize = 4.5,
  colCustom = keyvals,
  #col=c('#143D59', '#143D59', '#143D59', 'orange'),
  labCol = 'black',
  labFace = 'bold',
  boxedLabels = T,
  colAlpha = 3.5 / 5,
  legendPosition = 'top',
  legendLabSize = 18,
  legendIconSize = 5.0,
  drawConnectors = T,
  widthConnectors = 1.5,
  colConnectors = 'black',
  gridlines.major = T,
  gridlines.minor = FALSE,
  border = 'full',
  borderWidth = 1,
  borderColour = 'black',
  #xlim = c(-4,4),
  #ylim = c(0,11),
  selectLab = c("")
)

library(gridExtra)

grid.arrange(v1, v2, ncol=2)
    
```


## Differentially Regulated Genes - Visualized Interactively


```{r interactive, echo = F,fig.height = 6, fig.width = 10}

# DEGs Heatmap ------------------------------------------------------------

deseq_normalized_counts_degs_heatmap <- deseq_normalized_counts_degs %>%
  as.matrix() %>% 
  log2()

deseq_normalized_counts_degs_heatmap[is.infinite(deseq_normalized_counts_degs_heatmap)] <- 0

column_names <- colnames(deseq_normalized_counts_degs)

deseq_normalized_counts_degs_heatmap <-
  base::apply(deseq_normalized_counts_degs_heatmap, 1, scale) %>%
  t() 

colnames(deseq_normalized_counts_degs_heatmap) <- column_names

deseq_normalized_counts_degs_heatmap[is.nan(deseq_normalized_counts_degs_heatmap)] <- 0

colnames(deseq_normalized_counts_degs_heatmap) <- paste(meta$condition, meta$samples, sep = "-")

deseq_normalized_counts_degs_heatmap <-
  deseq_normalized_counts_degs_heatmap[order(rownames(deseq_normalized_counts_degs_heatmap)), order(colnames(deseq_normalized_counts_degs_heatmap))]

gradient_col <- ggplot2::scale_fill_gradient2(
  low = "blue", high = "red", 
  midpoint = 0
)

side_annot_top <- as.character(meta$condition) %>% 
  sort()

dit_cor_height = nrow(deseq_normalized_counts_degs_heatmap)*0.3
```

```{r heatmap_dit_cor, echo = F, fig.height=dit_cor_height, fig.width=10}
(heatmap_dit_cor <- heatmaply(
  deseq_normalized_counts_degs_heatmap,
  scale_fill_gradient_fun = gradient_col,
  limits = c(-1.5, 1.5),
  col_side_colors = data.frame("Condition" = side_annot_top),
  Rowv = TRUE,
  Colv = FALSE,
  showticklabels = TRUE,
  col_side_palette = colorRampPalette(c('#BF382A', '#0C4B8E'), bias=1),
  key.title = 'z-score'))

dit_cor_height2 = nrow(deseq_normalized_counts_degs_heatmap)*0.55
```

```{r heatmap_dit_cor2, echo=F, fig.height=dit_cor_height, fig.width=10}
(heatmap_dit_cor <- heatmaply(
  deseq_normalized_counts_degs_heatmap,
  scale_fill_gradient_fun = gradient_col,
  limits = c(-1.5, 1.5),
  col_side_colors = data.frame("Condition" = side_annot_top),
  Rowv = TRUE,
  Colv = T,
  showticklabels = TRUE,
  col_side_palette = colorRampPalette(c('#BF382A', '#0C4B8E'), bias=1),
  key.title = 'z-score'))


#Set annotation | https://www.biostars.org/p/317349/
ann <- data.frame(condition = meta$condition) %>%
  arrange(desc(condition))
colnames(ann) <- 'condition'
colours <-
  list('condition' = c('Asthma' = 'red2', 'NonAsthma' = 'royalblue'))
colAnn_s <- ComplexHeatmap::HeatmapAnnotation(
  df = ann,
  which = 'col',
  col = colours,
  annotation_width = unit(c(1, 4), 'cm'),
  gap = unit(1, 'mm')
)

```


```{r interactive2, echo=F, fig.height = 6, fig.width = 10}
# library(shiny)
# 
# col_fun2 = circlize::colorRamp2(c(-1.5, 0, 1.5),
#                                 c('mediumblue', 'white', 'red2'), space = 'sRGB')
# ht <-
#   ComplexHeatmap::Heatmap(
#     deseq_normalized_counts_degs_heatmap,
#     name = 'z-score',
#     col = col_fun2,
#     cluster_rows = T,
#     cluster_columns = F,
#     show_row_names = T,
#     border = F,
#     show_column_names = TRUE,
#     column_title = 'Samples sorted',
#     row_dend_width = unit(1, 'cm'),
#     heatmap_legend_param = list(legend_height = unit(6, "cm")),
#     top_annotation = colAnn_s,
#     column_names_gp = grid::gpar(fontsize = 7),
#     row_names_gp = grid::gpar(fontsize = 7)
#   )
# 
# # or use InteractiveComplexHeatmapWidget()
# ui = fluidPage(
#     actionButton("show_heatmap", "Generate_heatmap"),
#     htmlOutput("heatmap_output")
# )
# 
# server = function(input, output, session) {
#     observeEvent(input$show_heatmap, {
#         InteractiveComplexHeatmapWidget(input, output, session, draw(ht),
#             output_id = "heatmap_output")
#     })
# }
# shiny::shinyApp(ui, server, options = list(height = 650))
# 

```


```{r GSEA, echo = F, fig.width=15, fig.height=10}

## Top 30 most significant GO Terms - Visualized

# GSEA --------------------------------------------------------------------
#go_all_gsea_dge_out <- read.csv('exonic_GO_GSEA.csv')
#kegg_gsea_dge_out <- read.csv('exonic_KEGG_GSEA.csv')
#wiki_pathways_gsea <- read.csv('exonic_WIKIp_GSEA.csv')

# library(ggpubr)
# library(plotly)
# 
# (ggdc <- ggdotchart(
#   data.frame(go_all_gsea_dge_out)[1:30,],
#   x = 'Description',
#   y = "NES",
#   xlab = c(),
#   color = "ONTOLOGY",
#   # Color by groups
#   palette = c("#00AFBB", "#E7B800", "#FC4E07"),
#   # Custom color palette
#   sorting = "descending",
#   # Sort value in descending order
#   rotate = T,
#   add = "segments",
#   # Add segments from y = 0 to dots
#   add.params = list(color = "lightgray", size = 2),
#   # Change segment color and size
#   group = "ONTOLOGY",
#   # Order by groups
#   dot.size = 8,
#   # Large dot size
#   label = paste('', label = stringr::str_sub(go_all_gsea_dge_out$leading_edge[1:30], 6, 8)),
#   # Add mpg values as dot labels
#   font.label = list(
#     color = "white",
#     size = 9,
#     vjust = 0.5
#   ),
#   # Adjust label parameters
#   ggtheme = theme_pubr()                        # ggplot2 theme
# ) + geom_hline(yintercept = 0,
#                linetype = 2,
#                color = "lightgray")
# )
# 
# ggdc <- ggdotchart(
#   data.frame(go_all_gsea_dge_out)[1:30,],
#   x = 'Description',
#   y = "NES",
#   xlab = c(),
#   color = "ONTOLOGY",
#   # Color by groups
#   palette = c("#00AFBB", "#E7B800", "#FC4E07"),
#   # Custom color palette
#   sorting = "descending",
#   # Sort value in descending order
#   rotate = T,
#   add = "segments",
#   # Add segments from y = 0 to dots
#   add.params = list(color = "lightgray", size = 2),
#   # Change segment color and size
#   group = "ONTOLOGY",
#   # Order by groups
#   dot.size = 8,
#   # Large dot size
#   label = paste('', label = stringr::str_sub(go_all_gsea_dge_out$leading_edge[1:30], 6, 8)),
#   # Add mpg values as dot labels
#   font.label = list(
#     color = "white",
#     size = 9,
#     vjust = 0.5
#   ),
#   # Adjust label parameters
#   ggtheme = theme_pubr()                        # ggplot2 theme
# ) + geom_hline(yintercept = 0,
#                linetype = 2,
#                color = "lightgray") +
#     rremove('y.text') +
#   rremove('legend.title')
# 
# ggdc_kegg <- ggdotchart(
#   data.frame(kegg_gsea_dge_out)[1:30,],
#   x = 'Description',
#   y = "NES",
#   xlab = c(),
#   #color = "ONTOLOGY",
#   # Color by groups
#   palette = c("#00AFBB", "#E7B800", "#FC4E07"),
#   # Custom color palette
#   sorting = "descending",
#   # Sort value in descending order
#   rotate = T,
#   add = "segments",
#   # Add segments from y = 0 to dots
#   add.params = list(color = "lightgray", size = 2),
#   # Change segment color and size
#   #group = "ONTOLOGY",
#   # Order by groups
#   dot.size = 8,
#   # Large dot size
#   label = paste('', label = stringr::str_sub(go_all_gsea_dge_out$leading_edge[1:30], 6, 8)),
#   # Add mpg values as dot labels
#   font.label = list(
#     color = "white",
#     size = 9,
#     vjust = 0.5
#   ),
#   # Adjust label parameters
#   ggtheme = theme_pubr()                        # ggplot2 theme
# ) + geom_hline(yintercept = 0,
#                linetype = 2,
#                color = "lightgray") +
#     rremove('y.text')

```


## Top 30 most significant GO and KEGG Terms respectively - Visualized Interactively


```{r ggdc_plotly, echo = F, fig.height = 12, fig.width = 9}
#ggplotly(ggdc)



library(clusterProfiler)

# Data processing for GSEA Algorithm to GO and KEGG
dge_entrez <- bitr(
  dge$id,
  fromType = "ENSEMBL",
  toType = c("ENTREZID"),
  OrgDb = org.Hs.eg.db::org.Hs.eg.db
) %>%
  rename(entrez = ENTREZID,
         id = ENSEMBL) %>%
  full_join(dge)

geneList_dge <- dge_entrez$log2FoldChange
names(geneList_dge) <- dge_entrez$entrez
geneList_dge <- sort(geneList_dge, decreasing = TRUE)
geneList_dge_no_dups <-
  geneList_dge[!duplicated(names(geneList_dge))]
gene <- names(geneList_dge_no_dups)[abs(geneList_dge_no_dups) > 0.5]

dge_entrez_ora <- dge_entrez %>%
  filter(padj < 0.1)
geneList_dge_ora <- dge_entrez_ora$log2FoldChange
names(geneList_dge_ora) <- dge_entrez_ora$entrez
geneList_dge_ora <- sort(geneList_dge_ora, decreasing = TRUE)
geneList_dge_no_dups_ora <-
  geneList_dge_ora[!duplicated(names(geneList_dge_ora))]
gene_degs <-
  names(geneList_dge_no_dups_ora)[abs(geneList_dge_no_dups_ora) > 0.5]


the_seed = 1623250700 %% 100000 # 50700

# set the same seed for the following analysis using the GSEA algorithm
set.seed(the_seed)
go_all_gsea_dge_out <-
  clusterProfiler::gseGO(
    geneList     = geneList_dge_no_dups,
    OrgDb        = org.Hs.eg.db::org.Hs.eg.db,
    ont          = "all",
    pvalueCutoff = 1,
    verbose      = TRUE,
    seed         = TRUE,
    minGSSize    = 15,
    maxGSSize    = 500,
    keyType = 'ENTREZID'
  )

set.seed(the_seed)
kegg_gsea_dge_out <- gseKEGG(
  geneList     = geneList_dge_no_dups,
  organism     = 'hsa',
  minGSSize    = 15,
  maxGSSize    = 500,
  pvalueCutoff = 1,
  verbose      = TRUE,
  seed         = TRUE
)

set.seed(the_seed)
wiki_pathways_gsea <-
  gseWP(geneList_dge_no_dups, organism = "Homo sapiens")


#write.csv(data.frame(go_all_gsea_dge_out), file = '/Users/andrzejeljaszewicz/Erasmus_UP/liam_data_dge_analysis/featureCounts/exonic/exonic_GO_GSEA.csv')
#write.csv(data.frame(kegg_gsea_dge_out), file = '/Users/andrzejeljaszewicz/Erasmus_UP/liam_data_dge_analysis/featureCounts/exonic/exonic_KEGG_GSEA.csv')
#write.csv(data.frame(wiki_pathways_gsea), file = '/Users/andrzejeljaszewicz/Erasmus_UP/liam_data_dge_analysis/featureCounts/exonic/exonic_WIKIp_GSEA.csv')

library(DOSE)
library(ReactomePA)
set.seed(the_seed)
reactome_gsea <- gsePathway(
  geneList_dge_no_dups,
  pvalueCutoff = 1,
  pAdjustMethod = "BH",
  verbose = TRUE
)

set.seed(the_seed)
disease_ontology_do <- gseDO(
  geneList_dge_no_dups,
  minGSSize     = 15,
  pvalueCutoff  = 1,
  pAdjustMethod = "BH",
  verbose       = TRUE
)

set.seed(the_seed)
disease_ontology_ncg <- gseNCG(
  geneList_dge_no_dups,
  minGSSize     = 15,
  pvalueCutoff  = 1,
  pAdjustMethod = "BH",
  verbose       = TRUE
)

set.seed(the_seed)
disease_ontology_dgn <- gseDGN(
  geneList_dge_no_dups,
  minGSSize     = 15,
  pvalueCutoff  = 1,
  pAdjustMethod = "BH",
  verbose       = TRUE
)










library(ggpubr)
library(plotly)

# (ggdc <- ggdotchart(
#   data.frame(go_all_gsea_dge_out)[1:30,],
#   x = 'Description',
#   y = "NES",
#   xlab = c(),
#   color = "ONTOLOGY",
#   # Color by groups
#   palette = c("#00AFBB", "#E7B800", "#FC4E07"),
#   # Custom color palette
#   sorting = "descending",
#   # Sort value in descending order
#   rotate = T,
#   add = "segments",
#   # Add segments from y = 0 to dots
#   add.params = list(color = "lightgray", size = 2),
#   # Change segment color and size
#   group = "ONTOLOGY",
#   # Order by groups
#   dot.size = 8,
#   # Large dot size
#   label = paste('', label = stringr::str_sub(go_all_gsea_dge_out$leading_edge[1:30], 6, 8)),
#   # Add mpg values as dot labels
#   font.label = list(
#     color = "white",
#     size = 9,
#     vjust = 0.5
#   ),
#   # Adjust label parameters
#   ggtheme = theme_pubr()                        # ggplot2 theme
# ) + geom_hline(yintercept = 0,
#                linetype = 2,
#                color = "lightgray")
# )

ggdc <- ggdotchart(
  data.frame(go_all_gsea_dge_out)[1:30,],
  x = 'Description',
  y = "NES",
  xlab = c(),
  color = "ONTOLOGY",
  # Color by groups
  palette = c("#00AFBB", "#E7B800", "#FC4E07"),
  # Custom color palette
  sorting = "descending",
  # Sort value in descending order
  rotate = T,
  add = "segments",
  # Add segments from y = 0 to dots
  add.params = list(color = "lightgray", size = 2),
  # Change segment color and size
  group = "ONTOLOGY",
  # Order by groups
  dot.size = 8,
  # Large dot size
  label = paste('', label = stringr::str_sub(go_all_gsea_dge_out$leading_edge[1:30], 6, 8)),
  # Add mpg values as dot labels
  font.label = list(
    color = "white",
    size = 9,
    vjust = 0.5
  ),
  # Adjust label parameters
  ggtheme = theme_pubr()                        # ggplot2 theme
) + geom_hline(yintercept = 0,
               linetype = 2,
               color = "lightgray") +
    rremove('y.text') +
  rremove('legend.title')

ggdc_kegg <- ggdotchart(
  data.frame(kegg_gsea_dge_out)[1:30,],
  x = 'Description',
  y = "NES",
  xlab = c(),
  #color = "ONTOLOGY",
  # Color by groups
  palette = c("#00AFBB", "#E7B800", "#FC4E07"),
  # Custom color palette
  sorting = "descending",
  # Sort value in descending order
  rotate = T,
  add = "segments",
  # Add segments from y = 0 to dots
  add.params = list(color = "lightgray", size = 2),
  # Change segment color and size
  #group = "ONTOLOGY",
  # Order by groups
  dot.size = 8,
  # Large dot size
  label = paste('', label = stringr::str_sub(go_all_gsea_dge_out$leading_edge[1:30], 6, 8)),
  # Add mpg values as dot labels
  font.label = list(
    color = "white",
    size = 9,
    vjust = 0.5
  ),
  # Adjust label parameters
  ggtheme = theme_pubr()                        # ggplot2 theme
) + geom_hline(yintercept = 0,
               linetype = 2,
               color = "lightgray") +
    rremove('y.text')


p2<- subplot(ggdc, ggdc_kegg, margin = .075,titleY = TRUE, titleX=TRUE) 
p2

```
## Top 30 most significant GO Terms
```{r print_GO, echo=F}
knitr::kable((data.frame(go_all_gsea_dge_out))[1:30,2:10], format="markdown")
```
## Top 30 most significant KEGG Pathways
```{r print_KEGG, echo=F}
knitr::kable((data.frame(kegg_gsea_dge_out))[1:30,2:10], format="markdown")
```
## Top 30 most significant WIKI Pathways
```{r print_WIKIp, echo=F}
knitr::kable((data.frame(wiki_pathways_gsea))[1:nrow(data.frame(wiki_pathways_gsea)),2:10], format="markdown")
```
## Top 30 most significant Reactome Pathways
```{r print_reactome, echo=F}
knitr::kable((data.frame(reactome_gsea))[1:30,2:10], format="markdown")
```
## Top 30 most significant Disease Ontology - DO
```{r print_do_do, echo=F}
knitr::kable((data.frame(disease_ontology_do))[1:30,2:10], format="markdown")
```
## Top 30 most significant Disease Ontology - NCG
```{r print_do_cng, echo=F}
knitr::kable((data.frame(disease_ontology_ncg))[1:30,2:10], format="markdown")
```
## Top 30 most significant Disease Ontology - DGN
```{r print_do_dgn, echo=F}
knitr::kable((data.frame(disease_ontology_dgn))[1:30,2:10], format="markdown")
```
## Table of DEGs
```{r print_degs, echo=F}
degs_hgnc2 <-
  biomaRt::getBM(
    attributes = c('ensembl_gene_id', 'external_gene_name'),
    filters = 'ensembl_gene_id',
    values =  degs$id,
    mart = biomaRt::useDataset(
      "hsapiens_gene_ensembl",
      mart = biomaRt::useMart("ensembl", host = "asia.ensembl.org")
    )
  ) %>%
  dplyr::rename('id' = 'ensembl_gene_id',
         'hgnc' = 'external_gene_name') %>%
  full_join(degs)

knitr::kable(degs_hgnc2, format="markdown")
n_degs <- nrow(degs_hgnc2)
#knitr::kable(degs, format="markdown")
#n_degs <- nrow(degs)
```
## Number of DEGs: `r n_degs`

# COL + MMP + TIMP

```{r col, echo = F, fig.width=10, fig.height=9}

dge_hgnc <-
  biomaRt::getBM(
    attributes = c('ensembl_gene_id', 'external_gene_name'),
    filters = 'ensembl_gene_id',
    values =  dge$id,
    mart = biomaRt::useDataset(
      "hsapiens_gene_ensembl",
      mart = biomaRt::useMart("ensembl", host = "asia.ensembl.org")
    )
  ) %>%
  dplyr::rename('id' = 'ensembl_gene_id',
         'hgnc' = 'external_gene_name') %>%
  full_join(dge)

colmmptimp <- readxl::read_excel('col_mmp_timp.xlsx', sheet = 'all')

aim <- dge_hgnc %>% 
  filter(toupper(hgnc) %in% toupper(colmmptimp$hgnc))

knitr::kable(aim, format="markdown")


keyvals <-
  ifelse(aim$id %in% c(""),
         'red',
         ifelse((
           aim$log2FoldChange > 0.5 |
             aim$log2FoldChange < -0.5
         ) & aim$padj < 0.1,
         'orange',
         ifelse(aim$padj > 0.1, '#143D59', '#143D59')
         ))

keyvals[is.na(keyvals)] <- '#143D59'
#names(keyvals)[keyvals == 'red'] <- 'Commonly Regulated'
names(keyvals)[keyvals == '#143D59'] <-
  'Not Statistically Significant'
names(keyvals)[keyvals == 'orange'] <- 'Log2FC & padj'

v1 <- EnhancedVolcano::EnhancedVolcano(
  aim,
  lab = aim$hgnc,
  x = 'log2FoldChange',
  y = 'padj',
  title = 'Asthma vs Non-Asthma',
  pCutoff = 0.1,
  FCcutoff = 0.5,
  pointSize = 3.5,
  labSize = 4.5,
  colCustom = keyvals,
  #col=c('#143D59', '#143D59', '#143D59', 'orange'),
  labCol = 'black',
  labFace = 'bold',
  boxedLabels = T,
  colAlpha = 3.5 / 5,
  legendPosition = 'top',
  legendLabSize = 18,
  legendIconSize = 5.0,
  drawConnectors = T,
  widthConnectors = 1.5,
  colConnectors = 'black',
  gridlines.major = T,
  gridlines.minor = FALSE,
  border = 'full',
  borderWidth = 1,
  borderColour = 'black',
  #xlim = c(-4,4),
  #ylim = c(0,11)
)

v1

```