---
title: "RNA-Seq Differential Gene Expression and Quality Checks"
author: "Adrian Janucik"
date: "9/1/2021"
output: html_document
---

```{r Setup, include=FALSE}

#\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\#
#\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\#

normalized_matrix_file_name <- 'GTF_FIXED_deseq_featureCounts_normalised_counts.csv'

de_results_file_name <- 'GTF_FIXED_deseq_featureCounts_dge.csv'

meta_file_name <- 'sampleTable2.csv'

#\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\#
#\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\#

# Before deploying remember to add BiocManager to library sources
suppressPackageStartupMessages(library(BiocManager)) 

options(repos = BiocManager::repositories())

# Allow for overlaps on plots
options(ggrepel.max.overlaps = Inf)

# Dependencies 
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(InteractiveComplexHeatmap))
suppressPackageStartupMessages(library(dendextend))
suppressPackageStartupMessages(library(heatmaply))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(hrbrthemes))
suppressPackageStartupMessages(library(ggthemes))
suppressPackageStartupMessages(library(ggplot2))

# Set global options
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

```

## Quality Control And Differentiall Gene Expression Results

#### This Is A Report Of Quality Control And Differentail Gene Expression Analysis.

#### Differential Gene Expression Approach Used: DESeq2

#### Cutoffs To Establish Differentially Regulated Genes: |Log2FC| > 0.5 & padj < 0.1

## PCA Plots

```{r Load, echo=FALSE}

# Setup -------------------------------------------------------------------
# Read table with meta contents
meta <- read_csv(meta_file_name, 
                 show_col_types = FALSE)

# Specify factors and their levels
meta$condition <-
  factor(meta$condition, levels = c('NonAsthma', 'Asthma'))

# Create variables from meta
Condition <- meta$condition
Sample <- meta$samples

# Read in DESeq2 normalized count matrix (median of ratios)
normalized_counts <-
  read_csv(normalized_matrix_file_name,
           show_col_types = FALSE) %>%
  column_to_rownames(var = 'id') %>%
  select(-c(1)) # Depends

# Read in DE results
dge <- read_csv(de_results_file_name,
                show_col_types = FALSE) %>%
  select(-1) # Depends

```

```{r calculations_pca2d, echo=FALSE}
# PCA on DESeq2 normalized counts -----------------------------------------
# Perform PCA focusing only on first two dimensions for 2D plot
normalized_pca_2d <- normalized_counts %>%
  t() %>%
  prcomp(rank. = 2)

# Call summary on the PCA to acquire explained variance
normalized_pca_2d_sum <-
  summary(normalized_pca_2d)

# Create label for PCA showing % of explained variance
normalized_pca_2d_sum_total <-
  sum(normalized_pca_2d_sum$importance[2, 1:2]) * 100

```

##### 2D PCA on DESeq2's Normalized Counts

##### Total Explained Variance = `r normalized_pca_2d_sum_total`%

```{r plot_normalized_pca2d_all_i, echo=FALSE, fig.height=6, fig.width=9.5}

# Create PCA plot
normalized_pca2d_plot_all <- ggplot(data.frame(normalized_pca_2d$x), 
                                    aes(PC1, 
                                        PC2, 
                                        color = Condition, 
                                        label = Sample)) +
  geom_jitter(size = 3) +
  theme_hc()

# Convert PCA plot to be interactive 
(normalized_pca2d_plot_all_i <- normalized_pca2d_plot_all %>% 
    ggplotly() %>% 
    layout(margin=list(t = 75)))

```

```{r calculations_pca3d, echo = FALSE}
# Perform PCA focusing only on first two dimensions for 3D plot
normalized_pca_3d <- normalized_counts %>%
  t() %>%
  prcomp(rank. = 3)

# Call summary on the PCA to acquire explained variance
normalized_pca_3d_sum <-
  summary(normalized_pca_3d)

# Create label for PCA showing % of explained variance
normalized_pca_3d_sum_total <-
  sum(normalized_pca_3d_sum$importance[2, 1:3]) * 100

```

##### 3D PCA on DESeq2's Normalized Counts

##### Total Explained Variance = `r normalized_pca_3d_sum_total`%

```{r plot_normalized_pca3d_all_i, echo=FALSE, fig.height=6, fig.width=9.5}
(
  normalized_pca3d_plot_all_i <-
    plot_ly(
      data.frame(normalized_pca_3d$x),
      text = ~ rownames(normalized_pca_3d$x),
      x = ~ PC1,
      y = ~ PC2,
      z = ~ PC3,
      color = ~ Condition,
      colors = c('#BF382A', '#0C4B8E')
    ) %>%
    add_markers() %>%
    layout(scene = list(
      xaxis = list(title = paste(
        "PC1 -",
        round(normalized_pca_3d_sum$importance[2, 1], 3)
      )),
      yaxis = list(title = paste(
        "PC2 -",
        round(normalized_pca_3d_sum$importance[2, 2], 3)
      )),
      zaxis = list(title = paste(
        "PC3 -",
        round(normalized_pca_3d_sum$importance[2, 3], 3)
      ))
    ))
)
```

```{r calculations_pca2d_degs, echo = F}
# PCA on DESeq2 normalized counts - ONLY DEGs -----------------------------
# https://plotly.com/r/3d-scatter-plots/
# Filter DE results for DEGs
degs <- dge %>% 
  filter(padj < 0.1) %>% 
  filter(log2FoldChange > 0.5 | log2FoldChange < -0.5)

# Filter normalized matrix for DEGs
normalized_counts_degs <- normalized_counts[rownames(normalized_counts) %in% degs$id,]

# Perform 2D PCA using only DE genes on a 
normalized_pca_2d_degs <- normalized_counts_degs %>%
  t() %>%
  prcomp(rank. = 2)

# Call summary on the PCA to acquire explained variance
normalized_pca_2d_degs_sum <- summary(normalized_pca_2d_degs)

# Create label for PCA showing % of explained variance
tot_explained_variance_ratio_2d_degs <- sum(normalized_pca_2d_degs_sum$importance[2,1:2]) * 100

# Create 2D PCA plot
normalized_pca2d_plot_degs <- ggplot(data.frame(normalized_pca_2d_degs$x), 
                                     aes(PC1, 
                                         PC2, 
                                         color = Condition, 
                                         label = Sample)) +
  geom_jitter(size = 3) +
  theme_hc()

# Convert 2D PCA plot into interactive using plotly
normalized_pca2d_plot_degs_i <- normalized_pca2d_plot_degs %>% 
    ggplotly() %>% 
    layout(margin=list(t = 75))

```

##### 2D PCA on DESeq2's Normalized Counts

##### Total Explained Variance = `r tot_explained_variance_ratio_2d_degs`% 

```{r pca2d_2, echo = F, fig.height = 7, fig.width=9}
normalized_pca2d_plot_degs_i
```

```{r calculations_pca3d_degs, echo = FALSE}
# Perform PCA focusing only on first three dimensions
normalized_pca_3d_degs <- normalized_counts[rownames(normalized_counts) %in% degs$id,] %>%
  t() %>%
  prcomp(rank. = 3)

# Call summary on the PCA to acquire explained variance
normalized_pca_3d_degs_sum <-
  summary(normalized_pca_3d_degs)

# Create label for PCA showing % of explained variance
tot_explained_variance_ratio_3d_degs <-
  sum(normalized_pca_3d_degs_sum$importance[2, 1:3]) * 100

```

##### 3D PCA on DESeq2's Normalized Counts

##### Total Explained Variance = `r tot_explained_variance_ratio_3d_degs`%

```{r plot_normalized_pca3d_plot_degs, echo = FALSE, fig.height = 7, fig.width=9}
(normalized_pca3d_plot_degs <-
   plot_ly(
     data.frame(normalized_pca_3d_degs$x),
     text = ~rownames(normalized_pca_3d_degs$x),
     x = ~ PC1,
     y = ~ PC2,
     z = ~ PC3,
     color = ~ Condition,
     colors = c('#BF382A', '#0C4B8E')
   ) %>%
   add_markers() %>%
   layout(scene = list(
     xaxis = list(title = paste("PC1 -", round(normalized_pca_3d_degs_sum$importance[2, 1], 3))),
     yaxis = list(title = paste("PC2 -", round(normalized_pca_3d_degs_sum$importance[2, 2], 3))),
     zaxis = list(title = paste("PC3 -", round(normalized_pca_3d_degs_sum$importance[2, 3], 3)))
   )))

```

## Dendrograms representing sample clustering

##### Dendrogram on DESeq2's Normalized Counts

```{r dendrogram_all, echo = FALSE, fig.height = 7, fig.width = 9}
# Dendrogram on DESeq2 normalized counts  ---------------------------------
# Create distance matrix
normalized_counts_all_d <- dist(t(normalized_counts))
normalized_counts_all_d_h <- hclust(normalized_counts_all_d, method = "complete")
levelsCondition <- rev(levels(Condition))

dend_all <- as.dendrogram(normalized_counts_all_d_h)

# Color the branches based on the clusters:
dend_all_c <- color_branches(dend_all, k = 2, col = c('#BF382A', '#0C4B8E'))#, groupLabels = levelsCondition)

# Manually match the labels, as much as possible, to the real classification of the flowers:
labels_colors(dend_all_c) <-
  c('#BF382A', '#0C4B8E')[sort_levels_values(as.numeric(Condition)[order.dendrogram(dend_all_c)])]

# We shall add the flower type to the labels:
labels(dend_all_c) <-
  paste(as.character(Condition)[order.dendrogram(dend_all_c)],
        "(", labels(dend_all_c), ")",
        sep = "")

# We hang the dendrogram a bit:
dend_all_c_h <- hang.dendrogram(dend_all_c, hang_height = 0.1)

# reduce the size of the labels:
dend_all_c_h_s <- set(dend_all_c_h, "labels_cex", 0.6)

# And plot:
par(mar = c(2, 2, 2, 14) + 0.1, xpd = NA)

plot(dend_all_c_h_s,
     horiz =  TRUE,
     nodePar = list(cex = .007))

legend("topleft",
       legend = levelsCondition,
       fill = c('#BF382A', '#0C4B8E'))

colorCondition_all <-
  ifelse(as.character(Condition) == 'Asthma', '#BF382A', '#0C4B8E')

colored_bars(
  colors = colorCondition_all,
  dend = dend_all_c_h_s,
  rowLabels = "",
  add = TRUE,
  horiz = TRUE
)
```

#### Dendrogram on DESeq2 normalized counts with only DEGs

```{r dendrogram_degs, echo = FALSE, fig.height = 7, fig.width = 9}
# Dendrogram on DESeq2 normalized counts | DEGs only ---------------------
normalized_counts_degs <- normalized_counts[rownames(normalized_counts) %in% degs$id,]

# Create distance matrix
normalized_counts_degs_d <- dist(t(normalized_counts_degs))
normalized_counts_degs_d_h <- hclust(normalized_counts_degs_d, method = "complete")
levelsCondition <- rev(levels(Condition))

dend_degs <- as.dendrogram(normalized_counts_degs_d_h)

# Color the branches based on the clusters:
dend_degs_c <- color_branches(dend_degs, k = 2, col = c('#BF382A', '#0C4B8E'))#, groupLabels = levelsCondition)

# Manually match the labels, as much as possible, to the real classification of the flowers:
labels_colors(dend_degs_c) <-
  c('#BF382A', '#0C4B8E')[sort_levels_values(as.numeric(Condition)[order.dendrogram(dend_degs_c)])]

# We shall add the flower type to the labels:
labels(dend_degs_c) <-
  paste(as.character(Condition)[order.dendrogram(dend_degs_c)],
        "(", labels(dend_degs_c), ")",
        sep = "")

# We hang the dendrogram a bit:
dend_degs_c_h <- hang.dendrogram(dend_degs_c, hang_height = 0.1)

# reduce the size of the labels:
dend_degs_c_h_s <- set(dend_degs_c_h, "labels_cex", 0.6)

# And plot:
par(mar = c(2, 2, 2, 14) + 0.1, xpd = NA)

plot(dend_degs_c_h_s,
     horiz =  TRUE,
     nodePar = list(cex = .007))

legend("topleft",
       legend = levelsCondition,
       fill = c('#BF382A', '#0C4B8E'))

colorCondition_degs <-
  ifelse(as.character(Condition) == 'Asthma', '#BF382A', '#0C4B8E')

colored_bars(
  colors = colorCondition_degs,
  dend = dend_degs_c_h_s,
  rowLabels = "",
  add = TRUE,
  horiz = TRUE
)

```

## Correlation between samples performed on Euclidean distance with the use of DE genes

```{r calculations_heatmap_distance, echo = FALSE}
# Heatmap distance --------------------------------------------------------
# Heatmaply's gradient function is broken hence it will always use max and min matrix values as limits | ComplexHeatmap shiny app is better but as its an app markdown needs to be hosted in order to work
gradient_col <- ggplot2::scale_fill_gradient2(
  low = "blue", high = "red", 
  midpoint = 0, limits = c(-1, 1)
)

# Side annotations
side_annot <- meta$condition %>% 
  sort(decreasing = TRUE)

side_annot_top <- meta$condition %>% 
  sort(decreasing = FALSE)

# Input matrix
normalized_counts_d_r <- dist(t(normalized_counts),
                              diag = TRUE,
                              upper = TRUE) %>% 
  as.matrix() %>% 
  cor()

# Colnames and rownames pasted with its condition
colnames(normalized_counts_d_r) <- paste(meta$condition, meta$samples, sep = "-")
rownames(normalized_counts_d_r) <- colnames(normalized_counts_d_r)
sorted_normalized_counts_d_r <- normalized_counts_d_r[order(rownames(normalized_counts_d_r)), order(colnames(normalized_counts_d_r), decreasing = TRUE)]

# Relative height and width to the number of samples
normalized_counts_d_r_h = nrow(sorted_normalized_counts_d_r) * 0.25
normalized_counts_d_r_w = ncol(sorted_normalized_counts_d_r) * 0.35
```

```{r plot_heatmap_distance, echo = F, fig.height = normalized_counts_d_r_h, fig.width = normalized_counts_d_r_w}

# Construct the heatmap
normalized_counts_d_r_heatmap <- heatmaply(
  sorted_normalized_counts_d_r,
  scale_fill_gradient_fun = gradient_col,
  col_side_colors = data.frame("Condition" = side_annot_top),
  row_side_colors = data.frame("Condition" = side_annot),
  limits = c(-1,1),
  Rowv = FALSE,
  Colv = FALSE,
  showticklabels = TRUE,
  row_side_palette = colorRampPalette(c('#BF382A', '#0C4B8E'), bias=1),
  col_side_palette = colorRampPalette(c('#BF382A', '#0C4B8E'), bias=1),
  key.title = 'corr',
  colorbar_len = 0.01,
  side_color_colorbar_len = 0.01,
  subplot_heights = c(0.1,0.9),
  subplot_widths = c(0.9,0.1)) 

# Remove the duplicated labels caused by two sided annotations and remove the label
s <- subplot(normalized_counts_d_r_heatmap, margin = .01, titleY = TRUE) %>% 
  config(displayModeBar = FALSE)
s$x$data[[1]]$showlegend <- FALSE
s$x$data[[11]]$showlegend <- FALSE
s$x$layout$annotations[[2]]$text <- ''
s$x$layout$annotations[[2]]$y <- 0.8
s$x$layout$annotations[[2]]$x <- 1.026

s

```

## Volcano Plots

```{r volcanos, echo = FALSE, fig.height = 13, fig.width=18}
# Volcano -----------------------------------------------------------------
# Set color and labels according to Log2FC and padj
keyvals <-
  ifelse(dge$id %in% c(""),
         'red',
         ifelse((
           dge$log2FoldChange > 0.5 |
             dge$log2FoldChange < -0.5
         ) & dge$padj < 0.1,
         'orange',
         ifelse(dge$padj > 0.1, '#143D59', '#143D59')
         ))

keyvals[is.na(keyvals)] <- '#143D59'
#names(keyvals)[keyvals == 'red'] <- 'Commonly Regulated'
names(keyvals)[keyvals == '#143D59'] <-
  'Not Statistically Significant'
names(keyvals)[keyvals == 'orange'] <- 'Log2FC & padj'

# Construct volcano with labels

v1 <- EnhancedVolcano::EnhancedVolcano(
  dge,
  lab = dge$id,
  x = 'log2FoldChange',
  y = 'padj',
  title = 'Asthma vs Non-Asthma',
  pCutoff = 0.1,
  FCcutoff = 0.5,
  pointSize = 3.5,
  labSize = 4.5,
  colCustom = keyvals,
  #col=c('#143D59', '#143D59', '#143D59', 'orange'),
  labCol = 'black',
  labFace = 'bold',
  boxedLabels = T,
  colAlpha = 3.5 / 5,
  legendPosition = 'top',
  legendLabSize = 18,
  legendIconSize = 5.0,
  drawConnectors = T,
  widthConnectors = 1.5,
  colConnectors = 'black',
  gridlines.major = T,
  gridlines.minor = FALSE,
  border = 'full',
  borderWidth = 1,
  borderColour = 'black',
  #xlim = c(-4,4),
  #ylim = c(0,11)
)

# Construct volcano without labels
v2 <- EnhancedVolcano::EnhancedVolcano(
  dge,
  lab = dge$id,
  x = 'log2FoldChange',
  y = 'padj',
  title = 'Asthma vs Non-Asthma',
  pCutoff = 0.1,
  FCcutoff = 0.5,
  pointSize = 3.5,
  labSize = 4.5,
  colCustom = keyvals,
  #col=c('#143D59', '#143D59', '#143D59', 'orange'),
  labCol = 'black',
  labFace = 'bold',
  boxedLabels = T,
  colAlpha = 3.5 / 5,
  legendPosition = 'top',
  legendLabSize = 18,
  legendIconSize = 5.0,
  drawConnectors = T,
  widthConnectors = 1.5,
  colConnectors = 'black',
  gridlines.major = T,
  gridlines.minor = FALSE,
  border = 'full',
  borderWidth = 1,
  borderColour = 'black',
  #xlim = c(-4,4),
  #ylim = c(0,11),
  selectLab = c("")
)

# Plot both
library(gridExtra)
grid.arrange(v1, v2, ncol=2)

```

## Differentially Regulated Genes - Visualized Interactively

```{r interactive_DEGs, echo = FALSE}
# DEGs Heatmap ------------------------------------------------------------
# Construct the matrix with log2 transfored values to get the ~normalized distribution

normalized_counts_degs_heatmap <- normalized_counts_degs %>%
  as.matrix() %>% 
  log2()

# Make the initial values equal 0 to be 0 again
normalized_counts_degs_heatmap[is.infinite(normalized_counts_degs_heatmap)] <- 0

# Save the colnames
column_names <- colnames(normalized_counts_degs_heatmap)

# Calculate the z-score for a better visualisation
normalized_counts_degs_heatmap_z <-
  base::apply(normalized_counts_degs_heatmap, 1, scale) %>%
  t() 

# Bring back colnames
colnames(normalized_counts_degs_heatmap_z) <- column_names

# Make NaN values 0 again
normalized_counts_degs_heatmap_z[is.nan(normalized_counts_degs_heatmap_z)] <- 0

# Paste the conditions
colnames(normalized_counts_degs_heatmap_z) <- paste(meta$condition, meta$samples, sep = "-")

# Order the heatmap according to conditions
normalized_counts_degs_heatmap_z_o <-
  normalized_counts_degs_heatmap_z[order(rownames(normalized_counts_degs_heatmap_z)), order(colnames(normalized_counts_degs_heatmap_z))]

# Create the gradient | heatmaply limits are broken so it will only take into the account limits as max and min matrix values | ComplexHeatmap can fix it but it is a shiny app that makes the markdown needed to be hosted
gradient_col <- ggplot2::scale_fill_gradient2(
  low = "blue", high = "red", 
  midpoint = 0
)

# Sort annotations
side_annot_top <- as.character(meta$condition) %>% 
  sort()

# Fix the height and width
degs_height = nrow(normalized_counts_degs_heatmap_z_o)*0.2
degs_width = ncol(normalized_counts_degs_heatmap_z_o)*0.32
```

```{r heatmap_degs, echo = FALSE, fig.height = degs_height, fig.width = degs_width}
# DEGs heatmap with ROW clustering
(
  heatmaply(
    normalized_counts_degs_heatmap_z_o,
    scale_fill_gradient_fun = gradient_col,
    limits = c(-1.5, 1.5),
    col_side_colors = data.frame("Condition" = side_annot_top),
    Rowv = TRUE,
    Colv = FALSE,
    showticklabels = TRUE,
    col_side_palette = colorRampPalette(c('#BF382A', '#0C4B8E'), bias = 1),
    key.title = 'z-score',
    colorbar_len = 0.001,
    side_color_colorbar_len = 0.001,
    subplot_heights = c(0.05, 0.95),
    subplot_widths = c(0.95, 0.05)
  )
)

degs_height_clustered = nrow(normalized_counts_degs_heatmap_z_o) * 0.25
```

```{r heatmap_degs_clustering, echo = FALSE, fig.height = degs_height_clustered, fig.width = degs_width}
# DEGs heatmap with ROW and COL clustering
(
  heatmap_degs_hc <- heatmaply(
    normalized_counts_degs_heatmap_z_o,
    scale_fill_gradient_fun = gradient_col,
    limits = c(-1.5, 1.5),
    col_side_colors = data.frame("Condition" = side_annot_top),
    Rowv = TRUE,
    Colv = T,
    showticklabels = TRUE,
    col_side_palette = colorRampPalette(c('#BF382A', '#0C4B8E'), bias = 1),
    key.title = 'z-score',
    colorbar_len = 0.001,
    side_color_colorbar_len = 0.001,
    subplot_heights = c(0.05, 0.03, 0.92),
    subplot_widths = c(0.9, 0.1)
  )
)

```


```{r complex_heatmap, echo= FALSE, fig.height = 6, fig.width = 10}

# Commented out as its shiny app

# library(shiny)
# 
# col_fun2 = circlize::colorRamp2(c(-1.5, 0, 1.5),
#                                 c('mediumblue', 'white', 'red2'), space = 'sRGB')
# ht <-
#   ComplexHeatmap::Heatmap(
#     deseq_normalized_counts_degs_heatmap,
#     name = 'z-score',
#     col = col_fun2,
#     cluster_rows = T,
#     cluster_columns = F,
#     show_row_names = T,
#     border = F,
#     show_column_names = TRUE,
#     column_title = 'Samples sorted',
#     row_dend_width = unit(1, 'cm'),
#     heatmap_legend_param = list(legend_height = unit(6, "cm")),
#     top_annotation = colAnn_s,
#     column_names_gp = grid::gpar(fontsize = 7),
#     row_names_gp = grid::gpar(fontsize = 7)
#   )
# 
# # or use InteractiveComplexHeatmapWidget()
# ui = fluidPage(
#     actionButton("show_heatmap", "Generate_heatmap"),
#     htmlOutput("heatmap_output")
# )
# 
# server = function(input, output, session) {
#     observeEvent(input$show_heatmap, {
#         InteractiveComplexHeatmapWidget(input, output, session, draw(ht),
#             output_id = "heatmap_output")
#     })
# }
# shiny::shinyApp(ui, server, options = list(height = 650))
# 

```

## Top 30 most significant GO and KEGG Terms respectively - Visualized Interactively

```{r ggdc_plotly, echo = F, fig.height = 12, fig.width = 9}
# Data processing for GSEA Algorithm to GO and KEGG
library(clusterProfiler)

# Convert ENSEMBLID to ENTREZ
dge_entrez <- bitr(
  dge$id,
  fromType = "ENSEMBL",
  toType = c("ENTREZID"),
  OrgDb = org.Hs.eg.db::org.Hs.eg.db
) %>%
  rename(entrez = ENTREZID,
         id = ENSEMBL) %>%
  full_join(dge)

# Create geneList from the whole data set for GSEA
geneList_dge <- dge_entrez$log2FoldChange
names(geneList_dge) <- dge_entrez$entrez
geneList_dge <- sort(geneList_dge, decreasing = TRUE)
geneList_dge_no_dups <-
  geneList_dge[!duplicated(names(geneList_dge))]

# Also create gene list with genes passing the Log2FC
gene <- names(geneList_dge_no_dups)[abs(geneList_dge_no_dups) > 0.5]

# Also create gene list for DE genes
dge_entrez_ora <- dge_entrez %>%
  filter(padj < 0.1)
geneList_dge_ora <- dge_entrez_ora$log2FoldChange
names(geneList_dge_ora) <- dge_entrez_ora$entrez
geneList_dge_ora <- sort(geneList_dge_ora, decreasing = TRUE)
geneList_dge_no_dups_ora <-
  geneList_dge_ora[!duplicated(names(geneList_dge_ora))]
gene_degs <-
  names(geneList_dge_no_dups_ora)[abs(geneList_dge_no_dups_ora) > 0.5]

# After removing duplicates set seed for reproducibility
the_seed = 1623250700 %% 100000 # 50700

# set the same seed for the following analysis using the GSEA algorithm
# GENE ONTOLOGY
set.seed(the_seed)
go_all_gsea_dge_out <-
  clusterProfiler::gseGO(
    geneList     = geneList_dge_no_dups,
    OrgDb        = org.Hs.eg.db::org.Hs.eg.db,
    ont          = "all",
    pvalueCutoff = 1,
    verbose      = TRUE,
    seed         = TRUE,
    minGSSize    = 15,
    maxGSSize    = 500,
    keyType = 'ENTREZID'
  )

# KYOTO ENCYCLOPEDIA OF GENES AND GENOMES
set.seed(the_seed)
kegg_gsea_dge_out <- gseKEGG(
  geneList     = geneList_dge_no_dups,
  organism     = 'hsa',
  minGSSize    = 15,
  maxGSSize    = 500,
  pvalueCutoff = 1,
  verbose      = TRUE,
  seed         = TRUE
)

# WIKI PATHWAYS
set.seed(the_seed)
wiki_pathways_gsea <-
  gseWP(geneList_dge_no_dups, organism = "Homo sapiens")

# REACTOME
library(DOSE)
library(ReactomePA)
set.seed(the_seed)
reactome_gsea <- gsePathway(
  geneList_dge_no_dups,
  pvalueCutoff = 1,
  pAdjustMethod = "BH",
  verbose = TRUE
)

# DISEASE ONTOLOGY - disease ontology
set.seed(the_seed)
disease_ontology_do <- gseDO(
  geneList_dge_no_dups,
  minGSSize     = 15,
  pvalueCutoff  = 1,
  pAdjustMethod = "BH",
  verbose       = TRUE
)

# DISEASE ONTOLOGY - network of cancer gene
set.seed(the_seed)
disease_ontology_ncg <- gseNCG(
  geneList_dge_no_dups,
  minGSSize     = 15,
  pvalueCutoff  = 1,
  pAdjustMethod = "BH",
  verbose       = TRUE
)

# DISEASE ONTOLOGY - disease gene network
set.seed(the_seed)
disease_ontology_dgn <- gseDGN(
  geneList_dge_no_dups,
  minGSSize     = 15,
  pvalueCutoff  = 1,
  pAdjustMethod = "BH",
  verbose       = TRUE
)

# Generate lolipop charts
library(ggpubr)
library(plotly)
ggdc <- ggdotchart(
  data.frame(go_all_gsea_dge_out)[1:30,],
  x = 'Description',
  y = "NES",
  xlab = c(),
  color = "ONTOLOGY",
  # Color by groups
  palette = c("#00AFBB", "#E7B800", "#FC4E07"),
  # Custom color palette
  sorting = "descending",
  # Sort value in descending order
  rotate = T,
  add = "segments",
  # Add segments from y = 0 to dots
  add.params = list(color = "lightgray", size = 2),
  # Change segment color and size
  group = "ONTOLOGY",
  # Order by groups
  dot.size = 8,
  # Large dot size
  label = paste('', label = stringr::str_sub(go_all_gsea_dge_out$leading_edge[1:30], 6, 8)),
  # Add mpg values as dot labels
  font.label = list(
    color = "white",
    size = 9,
    vjust = 0.5
  ),
  # Adjust label parameters
  ggtheme = theme_pubr()                        # ggplot2 theme
) + geom_hline(yintercept = 0,
               linetype = 2,
               color = "lightgray") +
  rremove('y.text') +
  rremove('legend.title')

ggdc_kegg <- ggdotchart(
  data.frame(kegg_gsea_dge_out)[1:30,],
  x = 'Description',
  y = "NES",
  xlab = c(),
  #color = "ONTOLOGY",
  # Color by groups
  palette = c("#00AFBB", "#E7B800", "#FC4E07"),
  # Custom color palette
  sorting = "descending",
  # Sort value in descending order
  rotate = T,
  add = "segments",
  # Add segments from y = 0 to dots
  add.params = list(color = "lightgray", size = 2),
  # Change segment color and size
  #group = "ONTOLOGY",
  # Order by groups
  dot.size = 8,
  # Large dot size
  label = paste('', label = stringr::str_sub(go_all_gsea_dge_out$leading_edge[1:30], 6, 8)),
  # Add mpg values as dot labels
  font.label = list(
    color = "white",
    size = 9,
    vjust = 0.5
  ),
  # Adjust label parameters
  ggtheme = theme_pubr() # ggplot2 theme
) + geom_hline(yintercept = 0,
               linetype = 2,
               color = "lightgray") +
  rremove('y.text')

# Convert lolipops to interactive plots
subplot(ggdc, ggdc_kegg, margin = .075,titleY = TRUE, titleX=TRUE) 
```

## Top 30 most significant GO Terms

```{r print_GO, echo=F}
knitr::kable((data.frame(go_all_gsea_dge_out))[1:30,2:10], format="markdown")
```

## Top 30 most significant KEGG Pathways

```{r print_KEGG, echo=F}
knitr::kable((data.frame(kegg_gsea_dge_out))[1:30,2:10], format="markdown")
```

## Top 30 most significant WIKI Pathways

```{r print_WIKIp, echo=F}
knitr::kable((data.frame(wiki_pathways_gsea))[1:nrow(data.frame(wiki_pathways_gsea)),2:10], format="markdown")
```

## Top 30 most significant Reactome Pathways

```{r print_reactome, echo=F}
knitr::kable((data.frame(reactome_gsea))[1:30,2:10], format="markdown")
```

## Top 30 most significant Disease Ontology - DO - Disease Ontology

```{r print_do_do, echo=F}
knitr::kable((data.frame(disease_ontology_do))[1:30,2:10], format="markdown")
```

## Top 30 most significant Disease Ontology - NCG - Network of Cancer Gene

```{r print_do_cng, echo=F}
knitr::kable((data.frame(disease_ontology_ncg))[1:30,2:10], format="markdown")
```

## Top 30 most significant Disease Ontology - DGN - Disease Gene Network

```{r print_do_dgn, echo=F}
knitr::kable((data.frame(disease_ontology_dgn))[1:30,2:10], format="markdown")
```

## Table of DEGs

```{r print_degs, echo=F}
# Obtain HGNC symbols for DE genes
degs_hgnc2 <-
  biomaRt::getBM(
    attributes = c('ensembl_gene_id', 'external_gene_name'),
    filters = 'ensembl_gene_id',
    values =  degs$id,
    mart = biomaRt::useDataset(
      "hsapiens_gene_ensembl",
      mart = biomaRt::useMart("ensembl", host = "ensembl.org")
    )
  ) %>%
  dplyr::rename('id' = 'ensembl_gene_id',
                'hgnc' = 'external_gene_name') %>%
  full_join(degs)

knitr::kable(degs_hgnc2, format="markdown")
n_degs <- nrow(degs_hgnc2)
#knitr::kable(degs, format="markdown")
#n_degs <- nrow(degs)
```

## Number of DEGs: `r n_degs`

# COL + MMP + TIMP

```{r aim_of_study, echo = F, fig.width=10, fig.height=9}
# Fulfill the aim of the study
dge_hgnc <-
  biomaRt::getBM(
    attributes = c('ensembl_gene_id', 'external_gene_name'),
    filters = 'ensembl_gene_id',
    values =  dge$id,
    mart = biomaRt::useDataset(
      "hsapiens_gene_ensembl",
      mart = biomaRt::useMart("ensembl", host = "ensembl.org")
    )
  ) %>%
  dplyr::rename('id' = 'ensembl_gene_id',
                'hgnc' = 'external_gene_name') %>%
  full_join(dge)

colmmptimp <- readxl::read_excel('col_mmp_timp.xlsx', sheet = 'all')

aim <- dge_hgnc %>% 
  filter(toupper(hgnc) %in% toupper(colmmptimp$hgnc))

knitr::kable(aim, format="markdown")

keyvals <-
  ifelse(aim$id %in% c(""),
         'red',
         ifelse((
           aim$log2FoldChange > 0.5 |
             aim$log2FoldChange < -0.5
         ) & aim$padj < 0.1,
         'orange',
         ifelse(aim$padj > 0.1, '#143D59', '#143D59')
         ))

keyvals[is.na(keyvals)] <- '#143D59'
#names(keyvals)[keyvals == 'red'] <- 'Commonly Regulated'
names(keyvals)[keyvals == '#143D59'] <-
  'Not Statistically Significant'
names(keyvals)[keyvals == 'orange'] <- 'Log2FC & padj'

v3 <- EnhancedVolcano::EnhancedVolcano(
  aim,
  lab = aim$hgnc,
  x = 'log2FoldChange',
  y = 'padj',
  title = 'Asthma vs Non-Asthma',
  pCutoff = 0.1,
  FCcutoff = 0.5,
  pointSize = 3.5,
  labSize = 4.5,
  colCustom = keyvals,
  #col=c('#143D59', '#143D59', '#143D59', 'orange'),
  labCol = 'black',
  labFace = 'bold',
  boxedLabels = T,
  colAlpha = 3.5 / 5,
  legendPosition = 'top',
  legendLabSize = 18,
  legendIconSize = 5.0,
  drawConnectors = T,
  widthConnectors = 1.5,
  colConnectors = 'black',
  gridlines.major = T,
  gridlines.minor = FALSE,
  border = 'full',
  borderWidth = 1,
  borderColour = 'black',
  #xlim = c(-4,4),
  #ylim = c(0,11)
)

v3

```
